---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data
```{r Preprocessing, echo = TRUE}
options(scipen = 1, digits = 2)
# Checking the zip file is already there 
if (!file.exists("activity.zip")) {
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
method = "curl", destfile = "activity.zip", quiet = TRUE)
}
# Unzipping the zip file
unzip(zipfile = "activity.zip", exdir = "./")
# Reading the csv file
data <- read.csv("./activity.csv",header = TRUE)

```


## What is mean total number of steps taken per day?

```{r MeanStep, echo = TRUE}
library(plyr)
datanoNA <- na.omit(data)
stepsPerDay <- ddply(datanoNA,~date,summarise,sum(steps))
hist(stepsPerDay[,2],xlab ="Total number of steps taken per day",main = " ")
meanStepsPerDay <- mean(stepsPerDay[,2])
medianStepsPerDay <- median(stepsPerDay[,2])

```

- The mean total number of steps taken per day is `r as.numeric(meanStepsPerDay)`.
- The median total number of steps taken per day is `r as.numeric(medianStepsPerDay)`.





## What is the average daily activity pattern?
```{r AveragePattern, echo = TRUE}
n <- c(1:53)
stepsPerDay["newcolumn"] <- n
plot(stepsPerDay[,3],stepsPerDay[,2],type='l',ylab="Average Number of Steps per Day",xlab="Number of Days")
stepsPerDaywithNA <- ddply(data,~date,summarise,sum(steps))
totaldays <- nrow(stepsPerDaywithNA) 
sumStepswithNA <- sum(is.na(stepsPerDaywithNA))
```

Note that we do not have the original total of days (i.e, `r totaldays`) in our plot. Since we removed the missing values, in total `r sumStepswithNA` days, we ended up only with `r nrow(stepsPerDay)` days.


```{r Max, echo = TRUE}
max <- max(stepsPerDay[,2])
maximum <- stepsPerDay[which(stepsPerDay[,2]==max),]
```
The 5-minute interval number `r maximum[,3]`, on average across all the days in the dataset, contains the maximum number of steps. This maximum number of steps is `r maximum[,2]` and corresponds to the day `r maximum[,1]`.

## Imputing missing values

```{r ImputingMissingValues, echo = TRUE}
# Checking missing values
if (any(is.na(data))) {
sumNA <- sum(is.na(data))
colSums(is.na(data))
# Average of step in the dataset
averageSteps <- round(mean(data[,1],na.rm=TRUE))
# Imputing NA by the average of steps in the dataset
dataImputingNA <- data
dataImputingNA[which(is.na(dataImputingNA[,1])),] [,1] <- averageSteps
}

```
We have in our data `r sumNA` missing values.

Strategy to fill out the missing values using: 
Since the missing values appear all together on a particular day we can not use the mean/median for this day. Here I am imputing the missing values by the average of steps in the dataset. The new dataset with the missing data filled in is called "dataImputingNA".

```{r MeanStepImp, echo = TRUE}
stepsPerDayImp <- ddply(dataImputingNA,~date,summarise,sum(steps))
hist(stepsPerDayImp[,2],xlab ="Total number of steps taken per day",main = "Histogram Imputing Missing Values ")
meanStepsPerDayImp <- mean(stepsPerDayImp[,2])
medianStepsPerDayImp <- median(stepsPerDayImp[,2])

```

- The mean total number of steps taken per day after imputing missing values in the original dataset is `r as.numeric(meanStepsPerDayImp)`.
- The median total number of steps taken per day after imputing missing values in the original dataset is `r as.numeric(medianStepsPerDayImp)`.

Here we can compare with the mean and median using the original dataset removing the missing values:

- The mean total number of steps taken per day is `r as.numeric(meanStepsPerDay)`.
- The median total number of steps taken per day is `r as.numeric(medianStepsPerDay)`.

After comparing both values we can observe that they are different and for both cases, the mean and the median, imputing the missing value produces a decreasing in these values: the mean going from `r as.numeric(meanStepsPerDay)` to `r as.numeric(meanStepsPerDayImp)` and the median going from `r as.numeric(medianStepsPerDay)` to `r as.numeric(medianStepsPerDayImp)`.


## Are there differences in activity patterns between weekdays and weekends?
```{r DifferencesActPattern, echo = TRUE}
#Creating a new factor variable with two levels: "weekday" and "weekend"
# First I create a new dataset
dataImputingNAnew <- dataImputingNA
dataImputingNAnew$day <- weekdays(as.Date(dataImputingNAnew$date))
# I replace the weekend days by "weekend" word 
dataImputingNAnew$day[which(dataImputingNAnew$day == "Saturday" | dataImputingNAnew$day == "Sunday" )] <- "weekend"
# I Replace the weekdays by "weekday" word 
dataImputingNAnew$day[which(dataImputingNAnew$day == "Monday" | dataImputingNAnew$day == "Tuesday" | dataImputingNAnew$day == "Wednesday" | dataImputingNAnew$day == "Thursday" | dataImputingNAnew$day == "Friday" ) ] <- "weekday"
dataImputingNAnew$day <- as.factor(dataImputingNAnew$day)
# Generating the data for weekend day and weekday
numberWeekday <- dim(dataImputingNAnew[which(dataImputingNAnew$day =="weekend"),])[1]
numberWeekendDays <- dim(dataImputingNAnew[which(dataImputingNAnew$day =="weekday"),])[1]
dataImputingNAWeekend <- dataImputingNAnew[which(dataImputingNAnew$day =="weekend"),]
dataImputingNAWeekDays <- dataImputingNAnew[which(dataImputingNAnew$day =="weekday"),]
stepsPerDayImpWeekend <- ddply(dataImputingNAWeekend,~date,summarise,sum(steps))
stepsPerDayImpWeekday <- ddply(dataImputingNAWeekDays,~date,summarise,sum(steps))
nWeekend <- c(1:nrow(stepsPerDayImpWeekend))
nWeekday <- c(1:nrow(stepsPerDayImpWeekday))

stepsPerDayImpWeekend["newcolumn"] <- nWeekend
stepsPerDayImpWeekday["newcolumn"] <- nWeekday
# Plotting the data
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(stepsPerDayImpWeekend[,3],stepsPerDayImpWeekend[,2],type='l',ylab="Number of Steps",xlab="Number of Days",main= "weekend")
plot(stepsPerDayImpWeekday[,3],stepsPerDayImpWeekday[,2],type='l',ylab="Number of Steps",xlab="Number of Days",main= "weekday")

```


 Yes, from these plots we note that there are differences in activity patterns between weekdays and weekends. 

